Harnessing generative AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2024-11-15 manuscript.qmd
Edit manuscript

Ontologies provide a common language with which to communicate concepts. In medicine, ontologies for phenotypic abnormalities are invaluable for defining, diagnosing, prognosing, and treating human disease. Since 2008, the Human Phenotype Ontology (HPO) has been instrumental in healthcare and biomedical research by providing a framework for comprehensively describing human phenotypes and the relationships between them [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024]. By expanding its depth and breadth over time, the HPO now contains \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` diseases. Some HPO phenotypes also contain metadata annotations such typical age of onset, frequency, triggers, time course, mortality rate and typical severity. Describing the severity-related attributes of a disease is crucial for both research and clinical care of individuals with rare diseases. When researchers or clinicians are presented with phenotypes that fall outside of their expertise, resources to quickly and reliably retrieve summaries with additional relevant information about these phenotypes are essential. In the clinic, this can help in reaching a differential diagnosis or prioritising the treatment of some phenotypes over others. In research, this information is useful for prioritising targets for causal disease mechanisms, performing large-scale analyses of phenotypic data, and guiding funding agencies when assessing the potential impact and need for research in a given disease area. To date, the HPO has largely relied on manual curation by domain experts. While this approach can improve annotation quality and accuracy, it is both time-consuming and labour-intensive. As a result, less than 1% of terms within the HPO contain metadata such as time course and severity.

@@

plot_top_phenos_out <- HPOExplorer::plot_top_phenos(res_class = res_class)
plot_top_phenos_out$plot
```

@@

-   $NSS_p$: the final composite severity score for phenotype $p$ after applying normalisation to align values to a 0-100 scale and ensure equivalent meaning regardless of which other phenotypes are being analysed in addition to $p$. This allows for direct comparability of severity scores across studies with different sets of phenotypes.\

>>>>

Ontologies provide a common language with which to communicate concepts. In medicine, ontologies for phenotypic abnormalities are invaluable for defining, diagnosing, prognosing, and treating human disease. Since 2008, the Human Phenotype Ontology (HPO) has been instrumental in healthcare and biomedical research by providing a framework for comprehensively describing human phenotypes and the relationships between them [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024]. By expanding its depth and breadth over time, the HPO now contains \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities that are associated with \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` diseases as captured in HPOA [REF]. Some HPO phenotypes also contain information related to typical age of onset, frequency, triggers, time course, mortality rate and typical severity. Describing the severity-related attributes of a disease is crucial for both research and clinical care of individuals with rare diseases. When researchers or clinicians are presented with phenotypes that fall outside of their expertise, resources to quickly and reliably retrieve summaries with additional relevant information about these phenotypes are essential. In the clinic, this can help in reaching a differential diagnosis or prioritising the treatment of some phenotypes over others. In research, this information is useful for prioritising targets for causal disease mechanisms, performing large-scale analyses of phenotypic data, and guiding funding agencies when assessing the potential impact and need for research in a given disease area. To date, the HPO has largely relied on manual curation by domain experts. While this approach can improve annotation quality and accuracy, it is both time-consuming and labour-intensive. As a result, less than 1% of terms within the HPO contain metadata such as time course and severity.

@@

plot_top_phenos_out <- HPOExplorer::plot_top_phenos(res_class = res_class, axis.text.x=c(TRUE,TRUE))
plot_top_phenos_out$plot
```

@@

-   $NSS_p$: the final composite severity score for phenotype $p$ after applying normalisation to align values to a 0-100 scale and ensure equivalent meaning regardless of which other phenotypes are being analysed in addition to $p$. This allows for direct comparability of severity scores across studies with different sets of phenotypes.

{{< pagebreak >}
